---
title: "Disability1"
author: "Asher"
date: "2026-03-04"
output: html_document
---

library(tidyverse)
```{r}
setwd("~/Desktop/QSTUFF/")
library(tidyverse)
library(ggplot2)
library(forcats)
```


```{r}
acs_data <- read_csv("ACS_5-Year_Social_Characteristics_DC_Ward.csv")
View(acs_data)
```


Filtering for disability columns that start with DP02 codes (common in ACS)
```{r}
dp02_cols <- names(acs_data)[grepl("DP02", names(acs_data))]
print(dp02_cols)
```



Look specifically for disability-related columns (usually 0072-0081). This will show me exactly what disability columns are available
```{r}
disability_codes <- dp02_cols[grepl("007[2-9]|008[0-1]", dp02_cols)]
print(disability_codes)
```



Creating a data frame with descriptive names based on ACS standard definitions.
```{r}

disability_data <- data.frame(
  Disability_Type = c(
    "Total with Disability",
    "Hearing Difficulty",
    "Vision Difficulty",
    "Cognitive Difficulty",
    "Ambulatory Difficulty",
    "Self-Care Difficulty",
    "Independent Living Difficulty"
  ),
  
  Column = c(
    "DP02_0072E",  # Total civilian noninstitutionalized with disability
    "DP02_0073E",  # With a hearing difficulty
    "DP02_0074E",  # With a vision difficulty
    "DP02_0075E",  # With a cognitive difficulty
    "DP02_0076E",  # With an ambulatory difficulty
    "DP02_0077E",  # With a self-care difficulty
    "DP02_0078E"   # With an independent living difficulty
  ))

```



Forming the actual counts by summing across all wards, this creates a new column with the sum for each disability type. this chunk as remove any rows where the column wasn't found
```{r}
disability_data$Count <- sapply(disability_data$Column, function(col) {
  if(col %in% names(acs_data)) {
    sum(acs_data[[col]], na.rm = TRUE)
  } else {
    NA  
  }
})

disability_data <- disability_data[!is.na(disability_data$Count), ]

print(disability_data)
```



I calculated the total number of people in DC by age group, breaking them down into those with and without disabilities. I summed the counts from all wards for three age categories—under 18, 18 to 64, and 65 and over—and subtracted the disabled counts from the totals to get the non-disabled population. Then I printed the results in a simple table.
```{r}
dc_age_disability <- data.frame(
  Age_Group = c("Under 18", "18 to 64", "65 and Over"),
  Disabled = c(
    sum(social_data$DP02_0074E, na.rm = TRUE),  
    sum(social_data$DP02_0076E, na.rm = TRUE),  
    sum(social_data$DP02_0078E, na.rm = TRUE)   
  ),
  Non_Disabled = c(
    sum(social_data$DP02_0073E, na.rm = TRUE) - sum(social_data$DP02_0074E, na.rm = TRUE),   
    sum(social_data$DP02_0075E, na.rm = TRUE) - sum(social_data$DP02_0076E, na.rm = TRUE), 
    sum(social_data$DP02_0077E, na.rm = TRUE) - sum(social_data$DP02_0078E, na.rm = TRUE)   
  ))

print(dc_age_disability)
```



I created a bar graph showing the number of people with and without disabilities in each age group, using different colors for each status. By far the biggest disparity was between the 18 to 64 group, and the one witht he least was 65 and over with only 10K more non-disabled. 
```{r}
dc_age_disability <- data.frame(
  Age_Group = c("Under 18", "18 to 64", "65 and Over"),
  Disabled = c(11972, 40038, 26089),
  Non_Disabled = c(255680, 414344, 35633))

dc_age_disability$Age_Group <- factor(dc_age_disability$Age_Group, 
                                       levels = c("Under 18", "18 to 64", "65 and Over"))

dc_long <- dc_age_disability %>%
  pivot_longer(cols = c(Disabled, Non_Disabled),
               names_to = "Status",
               values_to = "Count") %>%
  mutate(Status = factor(Status, levels = c("Non_Disabled", "Disabled")))

ggplot(dc_long, aes(x = Age_Group, y = Count, fill = Status)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Non_Disabled" = "#E64B35", "Disabled" = "#4DBBD5"),
                    labels = c("Non_Disabled", "Disabled")) +
  labs(title = "Disability Status by Age Group in Washington, DC",
       subtitle = "Civilian Noninstitutionalized Population",
       x = "Age Group",
       y = "Population Count",
       fill = "Status",
       caption = "Source: ACS 5-Year Estimates (DP02), 2020-2024") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray30"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 11),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(labels = scales::comma, 
                     limits = c(0, 460000)) + 
  geom_text(aes(label = scales::comma(round(Count))), 
            position = position_dodge(width = 0.7),
            vjust = -0.5, 
            size = 3.5)
```


In percentage form, kept only the disabled percentages and printed them in a simple table
```{r}
percentages_output <- dc_long %>%
  group_by(Age_Group) %>%
  mutate(
    Total = sum(Count),
    Percentage = round((Count / Total) * 100, 1)
  ) %>%
  filter(Status == "Disabled") %>%
  select(Age_Group, Percentage_Disabled = Percentage)

print(percentages_output)
```



Loading in Curb ramo data and removing blanks and NAs
```{r}
ACR <- read.csv("ADA_Curb_Ramp.csv")
acr_counts <- ACR %>%
  filter(!is.na(CONDITION), CONDITION != "") %>%   
  count(CONDITION)
```



I analyzed the conditions of curb ramps by counting how many ramps fall into each condition category. I first calculated the overall average count to use as a reference line. Then, I created a plot showing each condition, with points sized by count and colored in a gradient. Then, I added a horizontal line for the average and segments connecting the average to each point to make differences easier to see
```{r}
overall_mean_cond <- ACR %>%
  filter(!is.na(CONDITION), CONDITION != "") %>%
  count(CONDITION) %>%
  summarise(mean_count = mean(n)) %>%
  pull(mean_count)

ACR_plot <- ACR %>%
  filter(!is.na(CONDITION), CONDITION != "") %>%
  count(CONDITION) %>%
  mutate(CONDITION = fct_reorder(CONDITION, n)) %>%
  ggplot(aes(x = CONDITION, y = n)) +
  
  labs(title = "Curb Ramp Condition Counts",
       x = "Condition",
       y = "Count") + 
  
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold")) +

  geom_hline(yintercept = overall_mean_cond, color = "lightblue", linewidth = 1) +
  geom_segment(aes(xend = CONDITION, y = overall_mean_cond, yend = n),
               color = "violet") +
  
  geom_point(aes(color = n), size = 4) +
  scale_color_gradient(low = "navy", high = "lightgreen")

ACR_plot
```

Loading Sidewalk data
```{r}
SCA <- read_csv("Sidewalk_Condition_Assessment.csv")
colnames(SCA)
```


I looked at the sidewalk condition data by counting how many sidewalks fall into each SCI category. I calculated the overall average count to add a reference line to the plot. Then, I made a plot with points for each category, colored by count, and added segments connecting the average line to each point to highlight differences.
```{r}
overall_mean_sci <- SCA %>%
  filter(!is.na(SCI_CATEGORY), SCI_CATEGORY != "") %>%
  count(SCI_CATEGORY) %>%
  summarise(mean_count = mean(n)) %>%
  pull(mean_count)

SCA_plot <- SCA %>%
  filter(!is.na(SCI_CATEGORY), SCI_CATEGORY != "") %>%
  count(SCI_CATEGORY) %>%
  mutate(SCI_CATEGORY = fct_reorder(SCI_CATEGORY, n)) %>%
  ggplot(aes(x = SCI_CATEGORY, y = n)) +
  
  labs(title = "Sidewalk Condition Category Counts",
       x = "SCI Category",
       y = "Count") + 
  
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold")) +
  
 
  geom_hline(yintercept = overall_mean_sci, color = "lightblue", linewidth = 1) +
  geom_segment(aes(xend = SCI_CATEGORY, y = overall_mean_sci, yend = n),
               color = "violet") +
  

  geom_point(aes(color = n), size = 4) +
  scale_color_gradient(low = "maroon", high = "orange")

SCA_plot
```



Total Sidewalk Lengths by Condition (SCI_CATEGORY)
```{r}
length_by_condition <- SCA %>%
  group_by(SCI_CATEGORY) %>%
  summarise(total_length = sum(SIDEWALK_LENGTH_FT, na.rm = TRUE)) %>%
  arrange(desc(total_length))

print(length_by_condition)
```



I removed any missing values for both the sidewalk condition category and the total length, then created a bar graph showing the total sidewalk length for each condition. I ordered the bars from greatest to least 
```{r}
length_by_condition %>%
  filter(!is.na(total_length), SCI_CATEGORY != "N/A") %>%
  mutate(SCI_CATEGORY = fct_reorder(SCI_CATEGORY, total_length, .desc = TRUE)) %>%
  ggplot(aes(x = SCI_CATEGORY, y = total_length)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Sidewalk Length by Condition",
    x = "Condition Category",
    y = "Total Length (ft)"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```



Percent of Sidewalks by Maintenance Priority
```{r}
priority_counts <- SCA %>%
  group_by(MAINTENANCEPRIORITY) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count) * 100)

print(priority_counts)
```


Cross-tab: CI_CATEGORY × MAINTENANCEPRIORITY
```{r}
crosstab_counts <- table(SCA$SCI_CATEGORY, SCA$MAINTENANCEPRIORITY)
print(crosstab_counts)
```


I made a stacked bar chart to see how maintenance priority is spread across different sidewalk conditions. Each bar shows the proportion of each priority level within a condition, so you can quickly tell which conditions have higher or lower maintenance needs.
```{r}
ggplot(SCA, aes(x = SCI_CATEGORY, fill = MAINTENANCEPRIORITY)) +
  geom_bar(position = "fill") +
  labs(
    title = "Maintenance Priority Distribution by Sidewalk Condition",
    x = "Condition Category",
    y = "Proportion",
    fill = "Maintenance Priority"
  ) +
  theme_minimal()
```



Length-weighted Cross-tab: This is better than counts because longer sidewalks matter more.
```{r}
length_crosstab <- SCA %>%
  group_by(SCI_CATEGORY, MAINTENANCEPRIORITY) %>%
  summarise(total_length = sum(SIDEWALK_LENGTH_FT, na.rm = TRUE))

print(length_crosstab)
```


Percentage of Fair, Good, Missing, Non-Compliant for Sidewalks
```{r}
ramp_percent <- curb %>%
  group_by(CONDITION) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count) * 100)

print(ramp_percent)
```

Check results (important)
```{r}
table(curb_clean$CONDITION, curb_clean$CONDITION_GROUP)
```


```{r}
curb_clean <- curb %>%
  filter(!is.na(CONDITION), CONDITION != "NA") %>%   # remove NA values
  mutate(CONDITION_GROUP = case_when(
    CONDITION %in% c("Good", "Fair") ~ "Compliant",
    CONDITION %in% c("Missing", "Non-Compliant") ~ "Non-compliant"
  ))
```


Calculate group percentages
```{r}
ramp_percent_grouped <- curb_clean %>%
  group_by(CONDITION_GROUP) %>%
  summarise(count = n()) %>%
  mutate(percent = count / sum(count) * 100)

print(ramp_percent_grouped)
```


